Test susieI using whole genome, s40.22 samples

### 2020-10-23 analysis/simulation-susieI-ukbWG-gtex.adipose_s40.22.Rmd
filtered samples s40.22

---
title: "Test susieI using whole genome, 4k ukb samples"
output: html_document
---

@@

```{r parameters}
n <- 40000 # number of samples
p <- 656321 # number of SNPs
J <- 8021 # number of genes
```

The genotype data we used is from UKB biobank, randomly selecting n = `r as.character(n)` samples. We use SNP genotype data from chr 1 to chr 22 combined from UKB. SNPs are downsampled to 1/10 (randomly), eQTLs (see below for definition of eQTL) were added back. This ends up with p = `p as.charater(p)` SNPs.

@@

* TWAS p value of SNPs:

@@

# Exploration of unusual small p values
Take chromosom 22 geno type and perform linear regression for against geno type for causal SNPs on other chromosomes. We see some very significant p values. Take chromosome 2 causal SNPs as an example.

## normal distribution as phenotype
```{r random, fig.height=4, fig.width=8}
chrom <- 22
load(paste0("~/causalTWAS/ukbiobank/ukb_chr", chrom ,"_s40000.FBM.Rd"))
load(paste0(simdatadir, "simu_s40000_GTEXadipose-B", chrom, "-cis-expr.Rd"))
load(paste0(simdatadir, "simu_", simtag, "-pheno.Rd"))

N<- nrow(phenores$Y)
cat("variance of the phenotype: ", var(phenores$Y), "\n")

snpname = dat$snp[,1]
anno <- cbind(dat$chr, dat$pos, dat$pos)

outname <- "/home/simingz/causalTWAS/simulations/simulation_susieI_20201001/chr22_plots/chr22snpRandom"
# GWAA(dat$G, matrix(rnorm(N, sd = sqrt(var(phenores$Y))), ncol =1), snpname = snpname, anno = anno, outname , family = gaussian, ncore = 5, nSplits = 30, compress = F)
n <- read.table(outname, header =T, comment.char = "")
plot(n$BEGIN, -log10(n$PVALUE), main = "Random", xlab = "position", ylab = "-log10(p)")
grid()
```

## genetic component of chromosome 2 as phenotype

Not all chromosome give small p values. chr2 is the one with the smallest p values.
```{r chrfixed, fig.height=4, fig.width=8}
chrom.p <- 2
r <- do.call(cbind, lapply(1:22, function(x) phenores$batch[[x]]$Y.g))
cat("variance of the genetic component: ", var(r[,chrom.p]))

outname <- "/home/simingz/causalTWAS/simulations/simulation_susieI_20201001/chr22_plots/chr22snpVschr2fixed"
# GWAA(dat$G, matrix(r[,chrom.p], ncol =1), snpname = snpname, anno = anno, outname, family = gaussian, ncore = 5, nSplits = 30, compress = F)
n <- read.table(outname, header =T, comment.char = "", stringsAsFactors = F)
plot(n$BEGIN, -log10(n$PVALUE), main = "chr2 genetic", xlab = "position", ylab = "-log10(p)")
grid()
```

## genotype of causal snps on chromosome 2 as phenotype

genotype of causal snps on chromosome 2 as phenotype. There are `r length(phenores$batch[[2]]$param$idx.cSNP)` causal SNPs. Not all of them significantly associated with SNP genotype on chr22, but quite a few. below are examples (only a subset of SNPs on chr22 are selected to run regression analysis with chr2 causal SNP genotypes.)

```{r geno, fig.height=4, fig.width=8, message=F}
chrom <- 22
load(paste0("~/causalTWAS/ukbiobank/ukb_chr", chrom ,"_s40000.FBM.Rd"))
geno <- dat$G[, match(n[n$PVALUE < 1e-40, "MARKER_ID"], dat$snp[,1])]
load(paste0("~/causalTWAS/ukbiobank/ukb_chr", chrom ,"_s40000.unscaled.FBM.Rd"))
geno.unscaled <-  dat$G[, match(n[n$PVALUE < 1e-40, "MARKER_ID"], dat$snp[,1])]

snpname = dat$snp[match(n[n$PVALUE < 1e-40, "MARKER_ID"], dat$snp[,1]),1]
anno <- cbind(dat$chr, dat$pos, dat$pos)[match(n[n$PVALUE < 1e-40, "MARKER_ID"], dat$snp[,1]), ]

load(paste0("~/causalTWAS/ukbiobank/ukb_chr", chrom.p ,"_s40000.FBM.Rd"))
geno.c <- dat$G[, phenores$batch[[chrom.p]]$param$idx.cSNP]
geno.c.names <- dat$snp[phenores$batch[[chrom.p]]$param$idx.cSNP,]
load(paste0("~/causalTWAS/ukbiobank/ukb_chr", chrom.p ,"_s40000.unscaled.FBM.Rd"))
geno.c.unscaled <- dat$G[, phenores$batch[[chrom.p]]$param$idx.cSNP]

ac <- ac <- apply(round(geno.unscaled), 2, table)
maf <- apply(cbind( (ac[1,] * 2  + ac[2,]) / N/2, 1- (ac[1,] * 2  + ac[2,]) /N/2), 1, min)

geno.c.select <- c(1, 23, 45, 65)

for (i in geno.c.select){
  outname <- paste0("/home/simingz/causalTWAS/simulations/simulation_susieI_20201001/chr22_plots/chr22snpVschr2causal", i)
  #GWAA(geno, matrix(geno.c[,i], ncol =1), snpname = snpname, anno = anno, outname, family = gaussian, ncore = 5, nSplits = 10, compress = F)
  rsq_max <- ld_max(geno.c.unscaled[,i, drop =F], geno.unscaled, stats = "R.squared")

  cat("max LD: ", rsq_max)
  m <- read.table(outname, header =T, comment.char = "")
  rbPal <- colorRampPalette(c('red','blue'))
  plot(anno[,2], -log10(m[m$X.CHROM ==22, ]$PVALUE),  main = paste0("chr", chrom.p, ":", geno.c.names[i]), xlab = "position", ylab = "-log10(p)", col =rbPal(11)[as.numeric(cut(maf,breaks = seq(0,0.5, 0.05)))], pch =20)
  legend("topleft",title="MAF",legend=seq(0,0.5, 0.05),col =rbPal(11),pch=20, cex = 0.5)
  grid()
}
```

@@

@@

>>>>

---
title: "Test susieI using whole genome, s40.22 samples"
output: html_document
---

@@

```{r parameters}
n.ori <- 40000 # number of samples
n <-  22542
p <- 656321 # number of SNPs
J <- 8021 # number of genes
```

The genotype data we used is from UKB biobank, randomly selecting `r as.character(n.ori)` samples. We then filtered samples based on relatedness, that ended up with n = `r as.character(n)` samples.  We use SNP genotype data from chr 1 to chr 22 combined from UKB. SNPs are downsampled to 1/10 (randomly), eQTLs (see below for definition of eQTL) were added back. This ends up with p = `p as.charater(p)` SNPs.

@@

*  p value of SNPs (GWAS):

@@

@@

## Regions: all (2)

* Susie run parameters: `L=2`. initialize with `null_weight = 0` and update `null_weight` based on last iteration results. We initialize prior for genes and SNPs using true values. gene.pve ~ 0.1, snp.pve ~ 0.5.

```{r param2, fig.height=6, fig.width=6}
tags <- paste(rep(1:2, each = 5), 1:5, sep = "-")
phenofs <- paste0(simdatadir, "simu_20201001-", tags, "-pheno.Rd")
susieIfs <- paste0(outputdir, "20201001-", tags, ".config2.susieIres.Rd")
susieIfs2 <- paste0(outputdir, "20201001-", tags, ".config2.susieI.txt")

@@

## Regions: filtered (1)
* We filterd regions with regional minimum FDR of both genes and SNPs < 0.05, round 700 regions left.
* Susie run parameters: `L=2`. initialize with `null_weight = 0` and update `null_weight` based on last iteration results. We initialize with prior for genes and SNPs as uniform. gene.pve ~ 0.1, snp.pve ~ 0.5.

```{r param3, fig.height=6, fig.width=6}
tags <- paste(rep(1:2, each = 5), 1:5, sep = "-")
phenofs <- paste0(simdatadir, "simu_20201001-", tags, "-pheno.Rd")
susieIfs <- paste0(outputdir, "20201001-", tags, ".config3.susieIres.Rd")
susieIfs2 <- paste0(outputdir, "20201001-", tags, ".config3.susieI.txt")

## Regions: filtered (2)
* We filterd regions with regional minimum FDR of both genes and SNPs < 0.01, round 400 regions left.
* Susie run parameters: `L=2`. initialize with `null_weight = 0` and update `null_weight` based on last iteration results. We initialize with prior for genes and SNPs as uniform. gene.pve ~ 0.1, snp.pve ~ 0.5.

```{r param4, fig.height=6, fig.width=6}
tags <- paste(rep(1:2, each = 5), 1:5, sep = "-")
phenofs <- paste0(simdatadir, "simu_20201001-", tags, "-pheno.Rd")
susieIfs <- paste0(outputdir, "20201001-", tags, ".config4.susieIres.Rd")
susieIfs2 <- paste0(outputdir, "20201001-", tags, ".config4.susieI.txt")
